---
title: "HW3"
author: "Anna Ma"
date: "2/24/2022"
output: pdf_document
---

```{r message = FALSE, warning=FALSE}
library(tidyverse)
```

### Problem 1

#### a)


1. Enter data

```{r}
cancer_alch_df = 
  tibble(age = rep(c(25, 35, 45, 55, 65, 75),2), 
         exposure = rep(c(0, 1), each = 6), 
         # exposure = 0: consumption 0-79g, exposure = 1: consumption >= 80g 
         case = c(0, 5,21,34,36,8,1,4,25,42,19,5), 
         control = c(106,164,138,139,88,31,9,26,29,27,18,0))
```

2. Model

```{r}
prosp_mod = glm(cbind(case, control) ~ age + exposure, data = cancer_alch_df, family = binomial(link = 'logit'))

prosp_mod %>% broom::tidy()
```


The fitted prospective model for the data is 
$$P(D = Disease |E = Exposure, X = Age) = \frac{e^{-5.02 + 0.0616X+1.78E}}{1+ e^{-5.02 + 0.0616X+1.78E}}$$

From the logit model, we can see that the log odds ratio of esophageal cancer between exposed (people with 0-79 g daily alcohol consumption) and unexposed group (people with 80+ g daily alcohol consumption) is `r exp(prosp_mod$coefficients[2])` for 1 unit change in age, given the same alcohol consumption

The log odds ratio of esophageal cancer between the exposed and unexposed group is `r exp(prosp_mod$coefficients[3])` for 1 unit change in daily alcohol consumption, given the same age.

Therefore, we can conclude that age and daily alcohol consumption of 80+ g are both positively associated with esophageal cancer.

#### b)

1. Fit both model

Let M0 be the smaller model where $\psi_j$,the odds ratio relating alcohol consumption and disease in the jth age group is 1; and M1 be the larger model where $\psi_j = \psi$

```{r}
cancer_alch_df = cancer_alch_df %>% 
  mutate(age_group = as.factor(c("1", "2", "3", "4", "5", "6", "1", "2", "3", "4", "5", "6")))

M0 = glm(cbind(case, control) ~ age_group, data = cancer_alch_df, family = binomial(link = 'logit'))

M1 = glm(cbind(case, control) ~ age_group+exposure, family = binomial(link = 'logit'),data = cancer_alch_df)

M0 %>% broom::tidy()
M1 %>% broom::tidy()
```

2. Check if the models are nested

```{r}
M0$coefficients
M1$coefficients
```

$M_0  = \beta_0 + \beta_1 * age_2 + \beta_2 * age_3 + \beta_3 * age_4 +\beta_4 * age_5 + \beta_5 * age_6$

M0 has 6 parameters and 5 predictors: $\beta_0,\beta_1,\beta_2,\beta_3,\beta_4,\beta_5$ & age_group 2-5

$M_1=\beta_0 + \beta_1 * age_2 + \beta_2 * age_3 + \beta_3 * age_4+ \beta_4 * age_5 + \beta_5 * age_6 + \beta_6 * alcohol$ 

M1 has 7 parameters and 6 predictors: $\beta_0,\beta_1,\beta_2,\beta_3,\beta_4,\beta_5,\beta_5$ & age_group 2-5 and alcohol consumption

Therefore, we can see that $M_0$ is nested in $M_1$, and we should perform deviance analysis.

Let $H_0: \beta_j = 0$, $H_a: \beta_j \neq 0$
```{r devi}
# Deviance
dev0 = M0$deviance #D1
dev1 = M1$deviance #D2
diff = dev0 - dev1 # difference between the two deviance
pchisq(diff, 6, lower.tail = FALSE) #p2 = the number of predictors in the larger model = 6
```

The difference of deviance between $M_0$ and $M_1$ is 79.52203, which gives a small p value of `r pchisq(diff, 6, lower.tail = FALSE)`. Therefore, we should abject the null hypothesis and conclude that $M_1$ fit the data better.


### Problem 2

#### a)

1. Enter Data

```{r}
germ_df = tibble(
  seed = c(rep("o75", 11), rep("o73", 10)), 
  root = c(rep("b", 5), rep("c", 6), rep("b", 5), rep("c", 5)),
  germ = c(c(10, 23, 23, 26, 17), c(5, 53, 55, 32, 46, 10), c(8, 10, 8, 23, 0), c(3, 22, 15, 32, 3)), 
  total = c(c(39, 62, 81, 51, 39), c(6, 74, 72, 51, 79, 13), c(16, 30, 28, 45, 4), c(12, 41, 30, 51, 7)))
```

2. Fit the logit regression model

```{r}
logit_mod = glm(cbind(germ, total-germ) ~ seed + root, data = germ_df, family = binomial(link = 'logit'))

logit_mod %>% broom::tidy()
```

Let G = Germinated, X1 = Seed, X2 = Root The fitted model is 
$$P(G | X_1 = x_1, X_2 = x_2) = \frac{e^{-0.7005 + 0.2705x_1 + 1.0647x_2}}{1+ e^{-0.7005 + 0.2705x_1 + 1.0647x_2}}$$
when $X_1 = 1$, the seed is O. aegyptiaca 75, and when $X_1 = 0$, the seed is O.aegyptiaca 73. When $X_2 = 1$, the root is cumcumber, and when $X_2 = 0$, the root is bean. 

```{r}
# intercept: germination rate for O73 seed in a bean root
exp(logit_mod$coefficients[1])
# b1: odds ratio of germination rate of O75 compare to O73, holding root constant yo be bean
exp(logit_mod$coefficients[2])
#b2: the odds ratio of germinating on a cucumber root
exp(logit_mod$coefficients[3])
```

From the model, we can see that the O.aegyptiaca 73 seed in bean root media has a `r exp(logit_mod$coefficients[1])` germination rate. 

Holding the root extract media constant to be bean, the odds ratio of germinating of O.aegyptiaca 75 seed is `r exp(logit_mod$coefficients[2])` time of an O.aegyptiaca 73 seed

Holding the seed type constant to be O.aegyptiaca 73, the odds ratio of germinating on a cucumber root is `r exp(logit_mod$coefficients[3])` time the odds ratio of a bean root. 

From those results, we can conclude that the root extract medium is highly associated with germinating rate.

#### b) dispersion

1. Goodness of fit
```{r}
logit_mod$deviance
qchisq(0.95,18)
```

The deviance of the model is `r logit_mod$deviance`, which is larger than the chi-square critical value of `r qchisq(0.95,18)` with 18 degrees of freedom. Therefore, the model is not a good fit.

2. Check for over dispersion with half normal Plot

```{r}
e = residuals(logit_mod, type = "pearson")
n_i = 1:21
res = tibble(
  x = qnorm((21 + n_i + 0.5)/(2 * 21 + 1.125)),
  y = sort(abs(e))
)

res %>% ggplot(aes(x = x,y = y)) +
  geom_point() +
  geom_smooth() +
  geom_abline(slope = 1) +
  labs(title = "Half Normal Plot",
       x = "Expected Half-Normal Order Statistics",
       y = "Ordered Pearson Residuals")
```

From the plot, we can see that there is a linear deviation from the reference line. Therefore, we can conclude that there is over dispersion.

3.Dispersion Parameter

```{r}
# pearson statistics
G = sum(residuals(logit_mod, type = "pearson")^2)
phi = G/(21 - 3)
```

The dispersion parameter $\phi$ = `r phi`

4. Update Model

```{r}
summary(logit_mod, dispersion = phi)
```

After updating the model, the estimate of the model did not change. However, the standard error was increased, which caused the p-value to change. In the updated model, we can clearly see that the p-value for seeds, 0.231 is greater than 0.05, indicating that the species of the seed might be insignificant to the germination rate, holding the root extract media constant.

#### c)

A plausible cause of the over dispersion is intra-class correlation. For example, germination in one spot can influence its neighbors, causing the germination to be dependent and violating the assumption that the trials are independent. 

